Staged Abstract Interpreters

### 2019-08-15 paper/50_stageabs.tex
Proofread 7

We have gradually presented the confluence of specialization and abstraction of
concrete interpreters. In this section, we review and summarize our recipe
to achieve the staged abstract interpreter, discuss the correctness issue
and different design choices.

@@

\begin{itemize}
  \item Constructing a generic interpreter that abstracts over binding-time,
    value domains and primitive operations. In this paper, the generic interpreter
    is implemented in monadic style, therefore the semantics can be encapsulated
    into monads.
  \item By using an approriate moand stack, we can implement a modular, unstaged
    abstract interpreter. This step has been explored in the previous literature.
  \item Replacing the monad stack to staged monad stack, such monads operate on
    staged data types, i.e., next-stage values. We also need to reimplement several
    primitive operations that on staged values, modularly.
\end{itemize}

Monadic interperters are known to be able to decouple the interpretation process
and the underlying semantics. The key insight in this paper is, by making the
monadic interpreter binding-time polymorphic, it can be extended for generating
efficient code in a staging framework. The underying semantics and staging are
two orthogonal dimensions. It is important to note that the computation
encapsulated by the monads are not staged, only data such as sets and maps are
staged. All the monadic computation, i.e., functions passed to the monadic bind
operation @flatMap@ are statically known. This is why we can eliminate the
monadic layer and its overhead.

@@

%\todo{may be show an small example?}

@@

As the central concern of static analysis, the soundness is important for
every user who would like to use it. Here we briefly justify that how our
approach preserve the corretness and soundness of the analysis, under the
assumption that the unstaged one is sound. It is worth noting that the rationale
listed here are based on empirical evidence, a rigorous proof of soundness
perservation of the staged abstract interpreters is still an open question.

\begin{itemize}
  \item We usually assume the underlying MSP system achieves cross-stage
    persistent, i.e., the generated program does not change the observational
    behaviours of the original program. In the case of this paper, we use the
    LMS framework, which 1) relies on the type checking facilities of Scala, and
    2) generates code in A-Normal form \cite{Flanagan:1993:ECC:155090.155113}
    that preserves the execution order within a stage \cite{DBLP:conf/birthday/Rompf16}.
    However, in general, it is possible to subvert these guarantees by
    using unsafe features, such as casts.
  \item As shown in the previous section, the generic interpreter is untouched
    and shared by the four artifacts. We only instantiate binding-time
    annotations and the monadic types. This allows the programmer to easily
    check the correctness of implementation of staged monads modularlly.
  \item We construct the staged monads and staged data structures in a
    correct-by-construct way that directly corresponds to their unstaged
    version. For instance, the staged data structure we used here are simply
    a black-box that wraps the Scala's library data structures (in the next
    stage).
  \item In our experiments, the staged abstract interpreter produces the same
    result with the unstaged one on all benchmark programs we tested.
\end{itemize}

@@

\paragraph{Are Monads Necessary?} No. One can always inline the monads and
obtain an abstract interpreter in continuation-passing style, or simply use
explicit side-effects such mutation to implement the artifact with the same
abstract semantics. In either case, we can still apply the staging schema to
the abstract interpreter. As evidence, we provide a staged abstract
interpreter written in direct-style in the accompanying artifact.

@@

What we implemented is a big-step, compositional abstract interpreter in monadic
style, where \textit{compositional} means that every recursive call of our abstract
interpreter is applied to proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as guarantees the termination
of specialization procedure. It is also possible to specialize small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587} -- as
\citet{Johnson:2013:OAA:2500365.2500604} implemented it for
optimizing Abstract Abstract Machines. However, the generated abstract
byte-code still requires another small-step abstract machine to execute, which is
an additional engineering effort.

>>>>

We have gradually presented the confluence of specialization and abstraction of
concrete interpreters. In this section, we review and summarize our recipe
to achieve the staged abstract interpreter, discuss correctness issues,
and different design choices.

@@

\begin{itemize}
  \item Constructing a generic interpreter that abstracts over binding-time,
    value domains, and primitive operations. In this paper, the generic interpreter
    is implemented in monadic style; therefore, the semantics can be encapsulated
    into monads.
  \item By using an appropriate monad stack, we can implement a modular, unstaged,
    abstract interpreter. This step has been explored in the previous literature.
  \item Replacing the monad stack to staged monad stack, such monads operate on
    staged data types, i.e., next-stage values. We also need to reimplement several
    primitive operations on staged values, in a modular fashion.
\end{itemize}

Monadic interpreters are known to be able to decouple the interpretation process
and the underlying semantics. The key insight in this paper is that by making the
monadic interpreter binding-time polymorphic, it can be extended for generating
efficient code in a staging framework. The underlying semantics and staging are
two orthogonal dimensions. It is important to note that the computation
encapsulated by the monads are not staged; only data (such as sets and maps) are
staged. All the monadic computation, i.e., functions passed to the monadic bind
operation @flatMap@, are statically known. This is why we can eliminate the
monadic layer and its associated overhead.

@@

%\todo{maybe show a small example?}

@@

Soundness is the central concern of static analysis, and as such, is vital
for prospective users. We now briefly examine how our
approach preserves the correctness and soundness of the analysis, under the
assumption that the unstaged one is sound. It is worth noting that the rationale
listed here is based on empirical evidence; a rigorous proof of soundness
preservation of the staged abstract interpreters remains an open question.

\begin{itemize}
  \item We usually assume the underlying MSP system achieves cross-stage
    persistence, i.e., the generated program does not change the observational
    behaviors of the original program. In the case of this paper, we use the
    LMS framework, which 1) relies on the type checking facilities of Scala, and
    2) generates code in A-Normal Form \cite{Flanagan:1993:ECC:155090.155113}
    that preserves the execution order within a stage \cite{DBLP:conf/birthday/Rompf16}.
    However, in general, it is possible to subvert these guarantees by
    using unsafe features, such as casts.
  \item As shown in the previous section, the generic interpreter is untouched
    and shared by the four artifacts. We only instantiate binding-time
    annotations and the monadic types. This allows the programmer to easily
    check the correctness of implementation of staged monads modularly.
  \item We construct the staged monads and staged data structures in a
    correct-by-construct way that directly corresponds to their unstaged
    version. For instance, the staged data structures we used here are simply
    black boxes that wrap the Scala's library data structures (in the next
    stage).
  \item In our experiments, the staged abstract interpreter produces the same
    result with the unstaged one on all benchmark programs we tested.
\end{itemize}

@@

\paragraph{Are Monads Necessary?} No. One can always inline the monads and
obtain an abstract interpreter in continuation-passing style, or simply use
explicit side-effects such as mutation to implement the artifact with the same
abstract semantics. In either case, we can still apply the staging schema to
the abstract interpreter. As evidence, we provide a staged abstract
interpreter written in direct-style in the accompanying artifact.

@@

We implemented a big-step, compositional abstract interpreter in monadic
style, where \textit{compositional} means that every recursive call of our abstract
interpreter is applied to proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as guarantees the termination
of the specialization procedure. It is also possible to specialize small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587} -- as
\citet{Johnson:2013:OAA:2500365.2500604} implemented it for
optimizing Abstract Abstract Machines. However, the generated abstract
bytecode still requires another small-step abstract machine to execute, which is
an additional engineering effort.
